Analysing Constraint Grammars with a SAT-solver

### 2016-02-12 paper/lrec/eval_futurework.tex
more text

\begin{itemize}
\item[] 
\begin{verbatim}SELECT DetPos IF (-1 (vbser pres p3 sg)) (0 ("zijn"i)) (1 Noun);
SELECT DetPosNotZijn IF (1 Noun);
\end{verbatim}
\end{itemize}

On the first glance it seems that there is no conflict: the first rule requires a Det Pos which is at the same time ``zijn'', as seen by the requirement \texttt{(0 ("zijn"i))}. However, this has a different semantics in the CG-3 formalism, as well as in SAT-CG. The rules match a situation where there is \emph{any} Det Pos and \emph{any} zijn, including the verb reading. The correct way would be

@@

We will work on the performance, and fix the errors we found in the
evaluation, such as the issue with non-exact position.
We also plan to restrict possible ambiguities in the symbolic words. 
Currently we have only hard-coded some
very basic assumptions, such as ``word cannot be ambiguous between a
boundary and a real word'', but we want to extract these constraints
automatically from a lexicon.
% For instance, the Spanish lexicon contains the entries
% \texttt{casa:casa<n><sg>}  and  \texttt{casa:casar<v><sg><p3>}, thus
% we know that a word can be ambiguous between a noun and a verb.
By adding these restrictions to the symbolic words, we hope to
restrict the search space, and make the SAT problem easier to solve, while at the same time generate more realistic example sentences.

>>>>

\begin{itemize}
\item[] 
\begin{verbatim}SELECT DetPos IF (-1 Verb) (0 "zijn") 
                 ( 1 Noun) ;
SELECT DetPosNotZijn IF (1 Noun) ;
\end{verbatim}
\end{itemize}

On the first glance it seems that there is no conflict: the first rule requires a Det Pos which is at the same time ``zijn'', as seen by the requirement \texttt{(0 "zijn")}. However, this has a different semantics in VISL CG-3, which is also repeated in our tool. The rules match a situation where there is \emph{any} Det Pos and \emph{any} zijn, including the verb. The correct way to formulate the first rule would be the following:

\begin{itemize}
\item[] \texttt{SELECT DetPos + "zijn" IF (-1 Verb) (1 Noun) ;}
\end{itemize}

@@

Our solution to hardcode the tag combinations in the readings is feasible for simple morphology, but it can cause problems with more complex morphology. One big downside is that in order to implement \textsc{ADD}, \textsc{ADDREADING} and \textsc{MAP}, we need to be prepared for new readings--even if the lexicon gives all readings that exist in the lexicon, the user might give a nonexistent reading, or in the case of MAP, a syntactic tag, which is (by definition) not in the lexicon. A more scalable solution would be to make each tag a variable, and ask the question ``can this reading be a noun? how about singular? how about conditional?'' separately for each tag. Then we could lift the restriction of tag combinations into a SAT solver: make SAT clauses that prohibit a comparative to go with a verb, or conditional with a noun.

### 2016-02-12 paper/lrec/implementation.tex
more text

Earlier we have shown an example with 5 readings (\texttt{det def}, \texttt{noun sg}, ...). In the real implementation, we operate between hundreds and thousands of readings. This is very much dependent on language: Dutch, the simplest language, has 336 readings.

We expand an Apertium morphological lexicon, throw away the word forms and take all the analyses. An exception is when the grammar contains rules that target a specific lemma or word form.
A simple solution is to retain the lemmas and word forms only for those entries where it is specified in the grammar, and otherwise leave them out. For example, the grammar may contain the following list:

\begin{itemize}
 \item[] \texttt{LIST ZijnVerb = ("zijn" vbser pres pl) ;}
\end{itemize}

Clearly there is something special about the verb \emph{zijn}, not just any verb. Looking at the lexicon, we find \emph{zijn} in the following entries:

@@

Another note: the readings in grammar can be underspecified (e.g. ``verb sg''), whereas the lexicon only gives us fully specified (``verb pres p2 sg'') readings.

@@

\todo{List of word forms+readings; list of distinct readings; make into matrix where we see that x and y can be ambiguous but a and z never; smart things with constraints (Koen)}

>>>>

Earlier we have shown an example with 5 readings (\texttt{det def}, \texttt{noun sg}, ...). In a realistic case, we operate between hundreds and thousands of readings. This is very much dependent on language: Dutch, the simplest language, has 336 readings.

We expand an Apertium morphological lexicon, throw away the word forms and take all the analyses. An exception is when the grammar contains rules that target a specific lemma or word form.
A simple solution is to retain the lemmas and word forms only for those entries where it is specified in the grammar, and otherwise leave them out. For example, the Dutch grammar contains the following rule:

\begin{itemize}
 \item[] \texttt{REMOVE ("zijn"i vbser) IF (-1 (pr)) (1 Noun) ;}
\end{itemize}

This hints that there is something special about the verb \emph{zijn}, compared to the other verbs. Looking at the lexicon, we find \emph{zijn} in the following entries:

@@

Another note: the readings in grammar can be underspecified (e.g. \texttt{verb sg}), whereas the lexicon only gives us fully specified (\texttt{verb pres p2 sg}) readings. We tried a version where we took the tag combinations specified in the grammar as readings, and we could insert them into the symbolic sentences as well, but this was not an ideal solution: turns out that the tag lists in the grammars contain often errors (e.g. replacing OR with an AND; using a nonexistent tag; using a wrong level in a subreading), and if we accept those lists as readings, we will generate symbolic sentences that are impossible, and won't discover the bug in the grammar.

However, if we only want to find rule interaction effects, then using the underspecified readings from the grammar makes the task faster, and it will still catch potential interaction errors.

@@

In the previous paragraph, we have created realistic \emph{readings}, by simply hardcoding legal tag combinations into variables. The next step in creating realistic ambiguities is to constrain what readings can go together. For instance, the case of \emph{zijn} shows us that ``determiner or verb'', is a possible ambiguity. In contrast, there is no word form in the lexicon that would be ambiguous between an adjective and a comma, hence we don't want to generate such ambiguity in our symbolic sentences.

We solve the problem creating \emph{ambiguity classes}: groups of analyses that can be ambiguous with each other. 
We represent the expanded morphological lexicon as a matrix, as seen in figure~\todo{make a nice picture with rows and columns}: word forms on the rows and analyses on the columns. Each distinct row forms an ambiguity class. For example, the ambiguity class that 
and form clusters based on which readings can be confused with each other. 
%This is done after the previous step: we simply treat the lemma or word form as another tag. 
Then we form SAT clauses that allow or prohibit certain combinations. These clauses will interact with the constraints created from the rules, and the end result will be closer to real-life sentences.

Our approach is similar to \todo{cite Cutting, Kupiec et al.}, who use ambiguity classes instead of distinct word forms, in order to reduce the number of parameters in a Hidden Markov Model. They take advantage of the fact that they don't have to model ``bear'' and ``wish'' as separate entries, but they can just reduce it to ``word that can be ambiguous between noun and verb'', and use it as a parameter in their HMM. 
We can do a similar thing by saving a list of words with each ambiguity class. For example, we map the ambiguity class ``feminine plural noun or a separable verb infinitive'' to the list of word forms \{``uitgaven'', ``toespraken''\}, and then, if we generate such reading for our symbolic word, we can give one of these words as an example word.
